---
title: "Database Loading and Extracting Baseball Data"
author: "Kris Eberwein"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Database Loading and Extracting Baseball Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The data contained in the Baseball Databank / Lahman Database are small enough to fit directly into memory using R. However, many may find it preferable to store the data in a local database to be accessed at a later time. The following examples use a local PostgreSQL database and the `RPostgreSQL` package. Many of the following methods would also work with other database connections such as the `RMySQL` and `DBI` packages.

## Loading Data

The goal here is to extract the data from source, do the transformations in R, and load them into the database. With the `baseballDBR` and `RPostgreSQL` packages, this is a simple task. The following assumes we are starting with an empty PostgreSQL instance named "lahman."

```{r, eval=FALSE}
library(baseballDBR)
library(RPostgreSQL)
library(dplyr)

get_bbdb(AllTables = TRUE)

# Make a list of all data frames.
dbTables <- names(Filter(isTRUE, eapply(.GlobalEnv, is.data.frame)))

# Load data base drivers and load all data frames in a loop.
drv <- dbDriver("PostgreSQL")
con <- dbConnect(drv, host= "localhost", dbname= "lahman", user= "YOUR_USERNAME", password = "YOUR_PASSWORD")

# Do some transformations on our selected dataframes.
Batting <- mutate(Batting, BA = BA(Batting), PA = PA(Batting), OBP = OBP(Batting), OPS = OPS(Batting))

# Create a loop to write all of our data frames our Postgres instance.
for (i in 1:length(dbTables)) { 
    if (dbExistsTable(con, dbTables[i])) {
       dbRemoveTable(con, dbTables[i])
    }
    dbWriteTable(con, name =  dbTables[i], value = get0(dbTables[i]), row.names = FALSE) 
}

# Make sure every thing wrote correctly.
test <- dbGetQuery(con, "SELECT * FROM Batting LIMIT 5")
test
rm(test)

# Close the database connection.
dbDisconnect(con)
dbUnloadDriver(drv)
```

## Accessing Data

Once our data are loaded into a local database instance, working with the data becomes less cumbersome because we have the choice to only load into R the data that we need, or the choice to leverage `dplyr` or `SQL` backends to work with the data without loading into R at all.

### Using dplyr Backends

The `dplyr` package allows us to work with remote or on-disk data stored in databases using a familiar syntax and `dplyr` verbs. For more on database backends, see the dplyr [Databases vignette](https://cran.r-project.org/package=dplyr). A current limitation to this method is, functions from the `baseballDBR` package will not work, and the method is limited to `dplyr` verbs.

```{r, eval=FALSE}
library(dplyr)
library(baseballDBR)

# Use an "src" connection to access the database. This requires the same arguments that RPostreSQL uses to connect to the database.
# Note: There are also src functions for connecting to various types of databases.
pgcon <- src_postgres(host= "localhost", dbname= "lahman", user= "YOUR_USERNAME", password = "YOUR_PASSWORD")

# We're not laoding the batting table, rather we're loading a method to access it.
Batting_tbl <- tbl(pgcon, "Batting")

# Use familiar dplyr verbs to access our data.
Batting_tbl %>% select(playerID, yearID, H, AB) %>% filter(yearID >= 1900)

# We can pull our data into R to permorm some modifications.
Batting <- collect(Batting_tbl, n = Inf) %>% mutate(BABIP = BABIP(Batting))

# Write the modified table back to the database.
db_insert_into( con = pgcon$con, table = "Batting", values = Batting)

```

### Using SQL Backends

For those familiar with SQL, the same process can be archived in our relational database's native language.

```{r, eval=FALSE}
library(RPostgreSQL)
library(baseballDBR)

# Load data base drivers and load all data frames in a loop.
drv <- dbDriver("PostgreSQL")
con <- dbConnect(drv, host= "localhost", dbname= "lahman", user= "YOUR_USERNAME", password = "YOUR_PASSWORD")

# Load the Batting table into R.
Batting <- dbGetQuery(con, "SELECT * FROM Batting")

# Permorm some modifications.
Batting <- mutate(BABIP = BABIP(Batting))

# Write modified table back to the database.
dbWriteTable("Batting", Batting)

# Close the database connection.
dbDisconnect(con)
dbUnloadDriver(drv)

```

